Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-04-05 Manuscript/manuscript.Rmd
manuscript edits

We assessed the collinearity among the genomic annotations used for the feature space of subsequent models by isolating the genomic bins that overlapped with flanked TAD boundaries. We then vectorized the number of instances when a region defined by a genomic annotation overlapped with the isolated bins. It was found that the correlations among these vectors were very high for many genomic annotations. The correlations were visualized using correlograms for each data resolution and cell line (Supplementary Figure 2). The high correlations imply that many genomic annotations tended to colocalize near each other at the boarders of domains. This was present across both resolution and cell line. Specifically, we found that the CTCF, Rad21, Smc3ab, and Znf143 transcription factor binding sites were highly colocalized near TAD boundaries. [??? We need to quantify and show this, and the following colocalizations] Likewise, many histone modifications also tended to appear near TAD boundaries in conjunction with one another including H3K4me1, H3K4me2, H3K4me3, H3K9ac, and H3K27ac. Furthermore, particular chromatin segmentation states, inclduing heterochromatin, repressed chromatin, poised promotors, and repetitive copy number variations, were found to be highly correlated, yet inversely correlated with many transcription factor binding sites.

The high correlations among many of the genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation. The regularization was performed after implementing each of the resampling techniques on all data resolutions within each cell line. Additionally, a unique predictor-type was included separately as the set of features. It was found that the elastic-net regularized the coefficients of far more [??? quantify] distance-type predictors to 0 compared to datasets with overlap count and percent predictor types, regardless of cell line or data resolution (Supplementary Figure 3). Many elastic-net models using overlap count or percent predictors were reduced to ridge, or near-ridge regression models at higher data resolutions by having the $alpha$ penalization term tuned to 0 in the cross-validation step, thereby not reducing, or reducing very few of the feature space. The results of the variable reduction step imply that far more information relating annotions to TAD boundaries is encoded in the distance-type predictors. That is, the elastic-net algorithms were better able to separate noise and reduce the feature space when performed on distance-type predictors. Likewise, this reduction was most efficient at more refined levels of data resolution.

## MODELS WITH DISTANCE-TYPE PREDICTORS OUTPERFORM MODELS WITH OTHER PREDICTOR-TYPES

@@

>>>>

By isolating bins that overlapped with flanked TAD boundaries we were able to assess the correlations of genomic annotations and determine which annotations were enriched together at boundaries. To  get a hollisitic view of how some annotations tended to occur together at boundaries, we computed the pairwise correlations of the averages between count overlaps, count percents, and $log_{2}$ distances of each genomic annotation with bins that overlapped with flanked TAD boundaries. Clustered correlograms at the 10 kb resolution, for both cell lines, showed that distinct groups of transcription factors were highly correlated including CTCF, ZNF143, RAD21, and SMC3ab (Figure 7). Likewise, combinations of particular histone modifications and chromatin states were also shown to exhibit similar profiles near TAD boundaries including H3K4me1, H3K4me2, and H3K4me3, aswell as strong and weak enhancers. Furthermore, particular chromatin segmentation states, inclduing heterochromatin, repressed chromatin, poised promotors, and repetitive copy number variations, were found to be highly correlated, yet inversely correlated with many transcription factor binding sites. Similar correlations were exhibited across the remainder of resolutions, for both cell lines (Supplementary Figure X). The high correlations among many of the genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation. The regularization was performed predictor-type and resolution specific data after implementing each of the resampling techniques within each cell line.

## CLASS IMBALANCE SEVERELY IMPACTS MODEL PERFORMANCE ACROSS RESOLUTIONS

@@

# LIMITATIONS

Many elastic-net models using overlap count or percent predictors were reduced to ridge, or near-ridge regression models at higher data resolutions by having the $alpha$ penalization term tuned to 0 in the cross-validation step, thereby not reducing, or reducing very few of the feature space. The results of the variable reduction step imply that far more information relating annotions to TAD boundaries is encoded in the distance-type predictors. That is, the elastic-net algorithms were better able to separate noise and reduce the feature space when performed on distance-type predictors. Likewise, this reduction was most efficient at more refined levels of data resolution.
